Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-21 papers/halfdisk/halfdisk.tex
Fixed typo
readthrough

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half disk) and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials.  We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations.
 \end{abstract}

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b, \qfor 0 \leq x \leq 1,\quad -\sqrt{1-x^2} \leq y \leq \sqrt{1-x^2},
$$
The corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

Sparsity comes from expanding the domain and range of an operator  using different choices of $a$ and $b$. Whereas the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle  results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

\sotodoinline{Mention that this is a computational challenge}

\sotodoinline{Make clear this is more than plugging in the basis into a Petrov--Galerkin method.}

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D.
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids \bstodo{Reference?}.

@@

\noindent \secref{OPs}: We present our procedure  to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the half disk domain, by combining 1D OPs to the interval, to form 2D OPs on the disk.

@@

\noindent\appref{PFEM}: We use the procedure to construct sparse $p$-finite element methods.

@@

By using a similar process to \cite[p55--56]{dunkl2014orthogonal} we can construct 2D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the intervals \([\alpha,\beta]\) and \([\gamma,\delta]\). We generalise this process to demonstrate a method of constructing multi-parameter families of 2D orthogonal polynomials on disk-slice and trapezium shaped domains $\Omega$ from multi-parameter families of 1D orthogonal polynomials on the intervals \([\alpha,\beta]\) and \([\gamma,\delta]\).

\begin{proposition}\label{prop:construction}
Let \(w_1 : (\alpha,\beta) \: \to \R\), \(w_2 : (\gamma,\delta) \: \to \R\) be weight functions with \(\alpha,\beta,\gamma,\delta \in \R\), and let \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w_2\) is an even function.
\end{enumerate}
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
$$
\hdopnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
$$
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

For disk slices and trapeziums, we can specialise the framework from Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{w_P^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: w_P^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: w_P^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Further, denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{P_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
We define the four-parameter 2D orthogonal polynomials on $\Omega$ by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
which are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
that is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}
\end{definition}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0,1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_P(x)$ is an ultraspherical weight, and the corresponding OPs are the Jacobi polynomials  $\{P_n^{(b, b)}\}$. The weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation $\{\genjac_n^{(a, b)}\}$ for the orthogonal polynomials. The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are orthonormal, with respect to the inner products respectively given by:
\begin{align*}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \\
	\ip<p, q>_{w_P^{(b)}} &:= \frac{1}{\normjac^{(b)}} \: \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}} = 1$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}} = 1$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that $\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}$ and that $\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}$.

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as 
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Hrecurrence} \\
	y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$:

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, since $\{\hdopmjab\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkab(x,) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjab(x,y)$. These coefficients are given by
\begin{align*}
	c_{m,j} = \frac{\ip< y \: \hdopnkab, \hdopmjab >_{\Wab}}{\norm{\hdopmjab}^2_{\Wab}}.
\end{align*}
Recall from equation (\ref{eqn:normhdop}) that $\norm{\hdopmjab}_{\Wab}^2 = \normgenjac^{(a,b+j+\half)} \: \normjac^{(b)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big)
\end{align*}
\begin{align*}
	=
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k+\frac{3}{2})} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k-\half)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}
\end{align*}
where $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}}$ and $\ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}}$ are zero for $m < n-1$.

@@

Recall that, for $a, b$ be general parameters, a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Recall also that $\Wab(x,y) := \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)})$, where $\rho(x) := (1-x^2)^\half$, $\genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b$ and $w_P^{(b)}(y) := (1-y^2)^b$ for the half disk case.

@@

\begin{lemma}\label{lemma:sparsityofdifferentialoperators}
The operator matrices $D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}$ from Definition \ref{def:differentialoperators} are sparse, with banded-block-banded structure.
\end{lemma}

@@

There exist "parameter transformation" matrix operators that increment/decrement the parameters, transforming the OPs from one (weighted or non-weighted) parameter space to another.

>>>>

\begin{abstract}
Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half disk) and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials.  We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations.
 \end{abstract}

@@

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b, \qfor 0 \leq x \leq 1,\quad -\sqrt{1-x^2} \leq y \leq \sqrt{1-x^2},
$$
The corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:construction}.

Sparsity comes from expanding the domain and range of an operator  using different choices of $a$ and $b$. Whereas the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle  results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside quadrature rules to determine the entries.  In particular, by exploiting the connection with one-dimensional orthogonal polynomials we are able to construct discretizations of general partial differential operators of size $p(p-1)/2 \times p(p-1)/2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This compares favourably to $O(p^6)$ operations if one proceeds na\"ively. Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

An alternative approach is to use \sotodoinline{Compare with "Discontinuous Galerkin scheme for the spherical shallow water equations with applications to tsunami modeling and prediction"}

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D.
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in turbulent pipe flow.

@@

\noindent \secref{OPs}: We present our procedure  to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the half disk domain, by combining 1D OPs on the interval, to form 2D OPs on the disk.

@@

\noindent\appref{PFEM}: We use the procedure to construct sparse $p$-finite element methods. This lays the groundwork for a future $hp$-finite element method in a disk, where the elements capture the circular geometry precisely.

@@

We can construct 2D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the intervals \([\alpha,\beta]\) and \([\gamma,\delta]\).

\begin{proposition}[{\cite[p55--56]{dunkl2014orthogonal}}]\label{prop:construction}
Let \(w_1 : (\alpha,\beta) \: \to \R\), \(w_2 : (\gamma,\delta) \: \to \R\) be weight functions with \(\alpha,\beta,\gamma,\delta \in \R\), and let \(\rho \: : \: (\alpha,\beta) \: \to (0,\infty)\) be such that either:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w_2\) is an even function.
\end{enumerate}
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then the 2D polynomials defined on $\Omega$
$$
\hdopnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)}) \qquad\hbox{for} \qquad 0 \le k \le n, \: n = 0,1,2,\dots
$$
are orthogonal polynomials with respect to the weight \(W(x,y) := w_1(x) \: w_2(\frac{y}{\rho(x)}) \) on $\Omega$. 
\end{proposition}

For disk slices and trapeziums, we  specialise  Proposition \ref{prop:construction} in the following definition.
\begin{definition}\label{def:construction}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
and define the associated inner products by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{w_P^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: w_P^{(a,b)}(y)\: \D y \label{eqn:ipjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: w_P^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
\end{align}
Further, denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{P_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = w^{(a,a)}_P(x) =  (\delta-x^2)^a$).
\end{enumerate}
We define the four-parameter 2D orthogonal polynomials on $\Omega$ by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
which are orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega,
\end{align*}
that is,
\begin{align*}
	\ip< \hdopnk^{(a,b,c,d)}, \: \hdopmj^{(a,b,c,d)} >_{W^{(a,b,c,d)}} &= \normgenjac^{(a,b,c+d+2k+1)} \: \normjac^{(d,c)} \: \delta_{n,m} \: \delta_{k,j},
\end{align*}
where for $f, g : \Omega \to \R$ the inner product is defined as 
\begin{align*}
	\ip< f, \: g >_{W^{(a,b,c,d)}} := \iint_\Omega \: f(x,y) \: g(x,y) \: W^{(a,b,c,d)}(x,y) \: \D y \: \D x.
\end{align*}
\end{definition}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0,1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of presentiation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_P(x)$ is an ultraspherical weight, and the corresponding OPs are the normalized Jacobi polynomials  $\{\tilde P_n^{(b, b)}\}$ \sotodo{Change $P_n -> \tilde P_n$ everywhere}. The weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation $\{\genjac_n^{(a, b)}\}$ for the orthogonal polynomials. The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are orthonormal, with respect to the inner products respectively given by:
\begin{align*}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \\
	\ip<p, q>_{w_P^{(b)}} &:= \frac{1}{\normjac^{(b)}} \: \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}} = 1$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}} = 1$ $\forall n = 0,1,2,\dots$. For clarity, we use the notation that $\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}$ and that $\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}$.

@@

We can express the three-term recurrences associated to $\genjac_n^{(a,b)}$ and $P_n^{(b,b)}$ as 
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Hrecurrence} \\
	y P_n^{(b,b)}(y) &= \delta_n^{(b)} P_{n+1}^{(b,b)}(y) + \gamma_n^{(b)} P_n^{(b,b)}(y) + \delta_{n-1}^{(b)} P_{n-1}^{(b,b)}(y),
\end{align}
where we note that \(\gamma_n^{(b)} = 0\), for \( n = 0,1,2,\dots\). We can use these to determine the 2D recurrences for $\hdopnkab(x,y)$. Importantly, we can deduce sparsity in the recurrence relationships:

@@

\begin{proof}
The 3-term recurrence for multiplication by $x$ follows from equation (\ref{eqn:Hrecurrence}). For the recurrence for multiplication by $y$, since $\{\hdopmjab\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n+1$ polynomial, we can expand $y \: \hdopnkab(x,) = \sum_{m=0}^{n+1} \sum_{j=0}^m c_{m,j} \: \hdopmjab(x,y)$. These coefficients are given by
\begin{align*}
	c_{m,j} = {\ip< y \: \hdopnkab, \hdopmjab >_{\Wab}}{\norm{\hdopmjab}^{-2}_{\Wab}}.
\end{align*}
Recall from equation (\ref{eqn:normhdop}) that $\norm{\hdopmjab}_{\Wab}^2 = \normgenjac^{(a,b+j+\half)} \: \normjac^{(b)}$. Then for $m = 0,\dots,n+1$, $j = 0,\dots,m$, using the change of variable $t = \frac{y}{\rho(x)}$:
\begin{align*}
	\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \rho(x)^{k+j+2} \: \genjacw^{(a, b)}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big) \\
	&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(x) \: \genjacmmj^{(a, b+j+\half)}(x) \: \genjacw^{(a, b+1+\half(k+j))}(x) \: \D x \Big) \nonumber \\
	& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: w_P^{(b)}(t) \: \D t \Big)
\end{align*}
\begin{align*}
	=
	\begin{cases}
    		\delta_k^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k+\frac{3}{2})} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
		\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \normgenjac^{(a,b+k-\half)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}} \quad& \text{if } j = k-1 \\
		0 & \text{otherwise}
      	\end{cases}
\end{align*}
where $\ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}}$ and $\ip<\genjacnmk^{(a, b+k+\half)}, \: \rho(x)^2 \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k-\half)}}$ are zero for $m < n-1$.

@@

Recall that a function $f(x,y)$ defined on $\Omega$ is approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Recall also that $\Wab(x,y) := \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)})$, where $\rho(x) := (1-x^2)^\half$, $\genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b$ and $w_P^{(b)}(y) := (1-y^2)^b$ for the half disk case.

@@

\begin{theorem}\label{theorem:sparsityofdifferentialoperators}
The operator matrices $D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}$ from Definition \ref{def:differentialoperators} are sparse, with banded-block-banded structure. \sotodoinline{Give block-bandwidths}
\end{theorem}

@@

There exist ``parameter transformation'' matrix operators that increment/decrement the parameters, transforming the OPs from one (weighted or non-weighted) parameter space to another.
